Trilemma

### 2022-04-14 Trilemma_clustering.Rmd
new figures for paper

```{r fig.width=6, fig.height=2}
df_summary <- df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    dplyr::summarize(sd_magnitude = sd(magni, na.rm = TRUE),
              sd_theta = sd(theta, na.rm = TRUE),
              sd_gamma = sd(gamma, na.rm = TRUE), 
              EF = mean(EFconsPerCap, na.rm = TRUE),
              Gini = mean(gini, na.rm = TRUE),
              GNI = mean(gni_log, na.rm = TRUE),
              sd_EF = sd(EFconsPerCap, na.rm = TRUE),
              sd_Gini = sd(gini, na.rm = TRUE),
              sd_GNI = sd(gni_log, na.rm = TRUE),
              #magnitude_m = mean(magni, na.rm = TRUE),
              #theta_m = mean(theta, na.rm = TRUE),
              #gamma_m = mean(gamma, na.rm = TRUE),
              magnitude_a = sum(magni, na.rm = TRUE),
              theta_a = sum(theta, na.rm = TRUE),
              gamma_a = sum(gamma, na.rm = TRUE)
              )

@@

pc1 <- ggbiplot::ggbiplot(
    pc, groups = as.character(stab@clusterObjs$kmeans$`3`$cluster),
    ellipse = TRUE, labels = df_summary$country_code, labels.size = 2.5,
    varname.size = 3) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    theme_light(base_size = 6) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank(),
          legend.text = element_text(size = 6), legend.title = element_text(size = 6),
          legend.key.width = unit(3,"mm"), legend.key.height = unit(2, "mm"))

@@

ggsave(
    filename = "pca_trilema.png", path = "figures/", device = "png",
    width = 4, height = 4, dpi = 300, bg = "white", 
    plot = pc1
)

@@

Averages are worth calculating for the parameter space, the vector field of the cube. But not for individual countries, or particules moving within the cube. Be aware of that when presenting and interpreting the results. 
```{r fig.height=3,fig.width=3}
a1 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(EF, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = EF, yend = Gini, color = as.factor(Group)),
        arrow = arrow(length = unit(0.1, "cm")), show.legend = FALSE, size = 0.15) +
    geom_text(
        aes(x = EF *1.1, y = Gini * 1.1, label = country_code, colour = Group), 
        size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(x = "Ecological Footprint", y = "Gini") +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2), 
          legend.text = element_text(size = 4), legend.title = element_text(size =4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

 ggplotly(c1, tooltip=c("country", "EFconsPerCap", "gini", "gni_log"))

@@

### Old stuff

Here is an attempt with another package `kml` that take into consideration the fact that our data is longitudinal. However, looking into the documentation, it seems that it only handles a variable at the time. For the sake of testing:

```{r}
# library(kml)

test_dat <- df_dat %>% 
  select(country, gini, year) %>% 
  as_tibble() %>% 
  pivot_wider(names_from = year, values_from = gini, names_prefix = "yr_") %>% 
  as_data_frame() %>% 
  kml::cld(idAll = "country", timeInData = 2:29)

```
An error I don't understand.

>>>>

```{r fig.width=6, fig.height=2}
df_summary <- df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    dplyr::summarize(sd_magnitude = sd(magni, na.rm = TRUE),
              sd_theta = sd(theta, na.rm = TRUE),
              sd_gamma = sd(gamma, na.rm = TRUE), 
              EF = mean(EFconsPerCap, na.rm = TRUE),
              Gini = mean(gini, na.rm = TRUE),
              GNI = mean(gni_log, na.rm = TRUE),
              sd_EF = sd(EFconsPerCap, na.rm = TRUE),
              sd_Gini = sd(gini, na.rm = TRUE),
              sd_GNI = sd(gni_log, na.rm = TRUE),
              #magnitude_m = mean(magni, na.rm = TRUE),
              #theta_m = mean(theta, na.rm = TRUE),
              #gamma_m = mean(gamma, na.rm = TRUE),
              magnitude = sum(magni, na.rm = TRUE),
              theta = sum(theta, na.rm = TRUE),
              gamma = sum(gamma, na.rm = TRUE)
              )

@@

pc1 <- ggbiplot::ggbiplot(
    pc, groups = as.character(stab@clusterObjs$kmeans$`3`$cluster),
    ellipse = TRUE, labels = df_summary$country_code, labels.size = 1.5,
    varname.size = 2) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(tag = "A") +
    theme_light(base_size = 6) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank(),
          legend.text = element_text(size = 6), legend.title = element_text(size = 6),
          legend.key.width = unit(3,"mm"), legend.key.height = unit(2, "mm"))

@@

ggsave(
    filename = "pca_trilema.png", path = "figures/", device = "png",
    width = 3, height = 3, dpi = 300, bg = "white",
    plot = pc1
)

@@

Averages are worth calculating for the parameter space, the vector field of the cube. But not for individual countries, or particules moving within the cube. Be aware of that when presenting and interpreting the results.

```{r fig.height=3,fig.width=3}
a1 <- df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    dplyr::summarize(
        EF = sum(d_EF, na.rm = TRUE),
        Gini = sum(d_gini, na.rm = TRUE),
        GNI = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(Group = as.character(stab@clusterObjs$kmeans$`3`$cluster)) |> 
    left_join(
        dat |> select(country, country_code) |> 
            filter(!is.na(country_code)) |> unique()) |> 
    ggplot(aes(EF, Gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = EF, yend = Gini, color = as.factor(Group)),
        arrow = arrow(length = unit(0.1, "cm")), show.legend = FALSE, size = 0.15) +
    geom_text(
        aes(x = EF *1.1, y = Gini * 1.1, label = country_code, colour = Group), 
        size = 1, show.legend = FALSE) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621","#5BA4CA")) +
    labs(x = "Ecological Footprint", y = "Gini") +
    theme_light(base_size = 6) +
    theme(panel.grid = element_blank(), legend.position = c(0.8,0.2), 
          legend.text = element_text(size = 4), legend.title = element_text(size =4),
          legend.key.width = unit(2,"mm"), legend.key.height = unit(2, "mm"))

@@

# ggplotly(c1, tooltip=c("country", "EFconsPerCap", "gini", "gni_log"))
f1
```

@@

## Static

```{r}
## ecological footprint
ef <- ecofp %>%
  filter(record == "EFConsPerCap") %>%
  select(countryName, isoa2, year, value) %>% 
  dplyr::rename(EFconsPerCap = value, country = countryName) |> 
    group_by(country, isoa2) |> 
    dplyr::summarize(
        m_ef = mean(EFconsPerCap, na.rm = TRUE), 
        sd_ef = sd(EFconsPerCap, na.rm = TRUE))

gini <- inq %>% 
    filter(!is.na(gini_reported)) %>% 
    select(country, isoa2 = c2, country_code = c3, year, gini_reported) %>% 
    group_by(country, year, isoa2, country_code) %>% 
    dplyr::summarize(gini_mean = mean(gini_reported, na.rm = TRUE)) |> 
    ungroup() |> group_by(country, isoa2, country_code) |> 
    dplyr::summarize(
        m_gini = mean(gini_mean, na.rm = TRUE),
        sd_gini = sd(gini_mean, na.rm = TRUE)
    )

gni <- dat1 %>% 
      select(-last_col()) %>% 
      filter(indicator_name == "GNI per capita (US$)") %>% 
      pivot_longer(cols = x1990:x2019, names_to = "year", values_to = "value") %>%
      mutate(year = str_remove(year, "x")) %>% 
      mutate(year = as.numeric(year)) %>% 
      mutate(ismissing = is.na(value)) %>% 
      dplyr::rename(country = country_name, gni = value) |> 
    group_by(country, country_code) |> 
    dplyr::summarize(
        m_gni = mean(gni, na.rm = TRUE),
        sd_gni = sd(gni, na.rm = TRUE)
    )

df_all <- gini |> left_join(ef) |> left_join(gni)

df_all <- df_all |> ungroup() |> 
    filter(!is.na(sd_gini), !is.na(m_ef), !is.na(sd_ef), !is.na(m_gni), !is.na(sd_gni)) 
```

```{r}
m <- "ward.D2" # minimize the total within-cluster variance

clust_num <- NbClust::NbClust( 
    data = (df_all %>% as.data.frame() %>% 
        select(where(is.numeric)) ),
    #distance = "maximum",
    min.nc = 2, max.nc = 7, 
    method = m, 
    index = 'all')
```

library(clValid)
library(mclust)
## Stability & Internal validation
stab <- clValid::clValid(
    obj = df_all %>% as.data.frame() %>% 
        select(where(is.numeric)) %>%
        as.matrix(),
    nClust = 2,
    clMethods = c(
        "hierarchical", "kmeans",  "som", 
        "model", "diana", "sota", "pam", "clara", "agnes"),
    validation = c('stability', "internal"),
    #metric = "manhattan",
    method = "ward",
    verbose = FALSE
) ## Hierarchical is the optimal method

clValid::optimalScores(stab)
```

```{r}
stab@clusterObjs$pam$`2`$clustering

```{r}
df_all |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering ) 
```

```{r}
pc_all <- prcomp(df_all |> 
                 select(where(is.numeric)) ,
             center = TRUE, scale. = TRUE)

pc2 <- ggbiplot::ggbiplot(
    pc_all, groups = as.character(stab@clusterObjs$pam$`2`$clustering ),
    ellipse = TRUE, labels = df_all$country_code, labels.size = 1.5,
    varname.size = 2) +
    scale_color_manual("Groups", values = c("#73B3A3","#FEA621")) +
    labs(tag = "B") + 
    theme_light(base_size = 6) +
    theme(legend.position = c(0.2, 0.2), panel.grid = element_blank(),
          legend.text = element_text(size = 6), legend.title = element_text(size = 6),
          legend.key.width = unit(3,"mm"), legend.key.height = unit(2, "mm"))

pc1 + pc2 + plot_layout(widths = 3, heights = 3)

ggsave(
    filename = "pca_trilemma.png", path = "figures/", device = "png",
    width = 6, height = 3, dpi = 300, bg = "white",
    plot = pc1+pc2
)
```

### Maps

```{r}
library(sf)
library(spData)
```

data(world)

ef_map <-  world |>
    filter(!is.na(pop)) |> 
    left_join(ef, by = c("iso_a2" = "isoa2")) |> 
    ggplot() +
    geom_sf(aes(fill = m_ef), size = 0.05, color = "white") +
    scico::scale_fill_scico(
        "Mean Ecological Footprint",palette = "romaO",
        guide = guide_colorbar(title.position = "top")) + 
    labs(tag = "A") +
    theme_void(base_size = 6) +
    theme(legend.position = "bottom", legend.key.height = unit(2, "mm"), 
          legend.key.width = unit(8, "mm"))

gini_map <-  world |>
    filter(!is.na(pop)) |> 
    left_join(gini, by = c("iso_a2" = "isoa2")) |> 
    ggplot() +
    geom_sf(aes(fill = m_gini), size = 0.05, color = "white") +
    scico::scale_fill_scico(
        "Mean Gini",palette = "berlin",
        guide = guide_colorbar(title.position = "top")) + 
    labs(tag = "B") +
    theme_void(base_size = 6) +
    theme(legend.position = "bottom", legend.key.height = unit(2, "mm"), 
          legend.key.width = unit(8, "mm"))

gni_map <-  world |>
    left_join(gini |> dplyr::select(isoa2, country_code),  
              by = c("iso_a2" = "isoa2") ) |> 
    left_join(gni |> ungroup() |> select(-country)) |> 
    filter(!is.na(pop)) |>
    ggplot() +
    geom_sf(aes(fill = m_gni), size = 0.05, color = "white") +
    scico::scale_fill_scico(
        "Mean Gross National Income",palette = "batlowW",
        guide = guide_colorbar(title.position = "top")) + 
    labs(tag = "C") +
    theme_void(base_size = 6) +
    theme(legend.position = "bottom", legend.key.height = grid::unit(2, "mm"), 
          legend.key.width = grid::unit(8, "mm"))

ef_map + gini_map + gni_map

ggsave(
    filename = "maps_trilemma.png", path = "figures/", device = "png",
    width = 6, height = 1.5, dpi = 400, bg = "white",
    plot = ef_map + gini_map + gni_map
)
